---
title: "Advent of Code 2021"
author: "Belay Birlie"
date: "01/12/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Day 1: Sonar Sweep

The puzzle for Day - 1 of the [advent of code](https://adventofcode.com/2021/) for 2021 is to count the number of times a given measurement increases from the previous measurement. The measurents are 1 dimenional vector.  
```{r housekeeping}

library(here)
here::i_am("Solution/Adventofcode.Rmd")

```

Let's first read the data. I saved the text file in the *data* folder and package *here* makes it easy for me to find the the data. I will use the base function diff with the defult parameter setting, that is lag = 1 

```{r day1_puzzle}
mmeasurement<-as.integer(readLines(here("data/measurement_day1.txt")))
sum(diff(mmeasurement)>0)

```

The second part of the puzzle consider sums of a three-measurement sliding window and the goal is to count the number of times the sum of measurements in this sliding window increases from the previous sum. I will use *zoo* package to solve this 

```{r day1_puzzle2}
sum(diff(zoo::rollapply(mmeasurement, 3, sum))>0)

```
