---
title: "Advent of Code 2021"
author: "Belay Birlie"
date: "01/12/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Day 1: Sonar Sweep

The puzzle for Day - 1 of the [advent of code](https://adventofcode.com/2021/) for 2021 is to count the number of times a given measurement increases from the previous measurement. The measurents are 1 dimenional vector.  
```{r housekeeping}

library(here)
here::i_am("Solution/Adventofcode.Rmd")

```

Let's first read the data. I saved the text file in the *data* folder and package *here* makes it easy for me to find the the data. I will use the base function diff with the defult parameter setting, that is lag = 1 

```{r day1_puzzle}
mmeasurement<-as.integer(readLines(here("data/measurement_day1.txt")))
sum(diff(mmeasurement)>0)

```

The second part of the puzzle consider sums of a three-measurement sliding window and the goal is to count the number of times the sum of measurements in this sliding window increases from the previous sum. I will use *zoo* package to solve this 

```{r day1_puzzle2}
sum(diff(zoo::rollapply(mmeasurement, 3, sum))>0)

```

## Day 2

The submarine seems to already have a planned course (your puzzle input). You should probably figure out where it's going. For example:

  - forward 5
  - down 5
  - forward 8
  - up 3
  - down 8
  - forward 2

Your horizontal position and depth both start at 0. The steps above would then modify them as follows:

forward 5 adds 5 to your horizontal position, a total of 5.
down 5 adds 5 to your depth, resulting in a value of 5.
forward 8 adds 8 to your horizontal position, a total of 13.
up 3 decreases your depth by 3, resulting in a value of 2.
down 8 adds 8 to your depth, resulting in a value of 10.
forward 2 adds 2 to your horizontal position, a total of 15.
After following these instructions, you would have a horizontal position of 15 and a depth of 10. (Multiplying these together produces 150.)

Calculate the horizontal position and depth you would have after following the planned course. What do you get if you multiply your final horizontal position by your final depth?

```{r day2_puzzle2}

library(data.table)
day2<-readLines(here("data/day2.txt"))
day2<-as.data.frame(setDT(as.data.frame(day2))[, paste0("v", 1:2) := tstrsplit(day2, " ")][, .(sum.depth = sum(as.numeric(v2))), by = v1])
(day2$sum.depth[day2$v1 == "down"] - day2$sum.depth[day2$v1 == "up"])*day2$sum.depth[day2$v1 == "forward"]


```


## Day 3 puzzle part 1

iven the following diagnostic report:

00100
11110
10110
10111
10101
01111
00111
11100
10000
11001
00010
01010
Considering only the first bit of each number, there are five 0 bits and seven 1 bits. Since the most common bit is 1, the first bit of the gamma rate is 1.

The most common second bit of the numbers in the diagnostic report is 0, so the second bit of the gamma rate is 0.

The most common value of the third, fourth, and fifth bits are 1, 1, and 0, respectively, and so the final three bits of the gamma rate are 110.

So, the gamma rate is the binary number 10110, or 22 in decimal.

The epsilon rate is calculated in a similar way; rather than use the most common bit, the least common bit from each position is used. So, the epsilon rate is 01001, or 9 in decimal. Multiplying the gamma rate (22) by the epsilon rate (9) produces the power consumption, 198.

Use the binary numbers in your diagnostic report to calculate the gamma rate and epsilon rate, then multiply them together. What is the power consumption of the submarine? (Be sure to represent your answer in decimal, not binary.)

```{r day3_puzzle1}
day3<-readLines(here("data/day3.txt"))
day3<-do.call( "rbind", strsplit(day3, split = ""))
day3 <- matrix(as.numeric(day3), ncol = 12)
strtoi(paste(ifelse(colSums(day3)/nrow(day3) >0.5, 1,0),collapse=""),base=2)*
  strtoi(paste(abs(ifelse(colSums(day3)/nrow(day3) >0.5, 1,0)-1),collapse=""),base=2)
  
```

